Viscous Erosion in Porous Media

### 2018-03-06 docs/TexNotes/Notes.tex
Working on drag. It is kind of subtle to get estimates with slender-body theory.

\section{Scaling law and interface smoothing}
For 3D Stokes flow, the shear stress scales like
\begin{equation}
\tau \sim \mu V/L
\end{equation}
where $V$ and $L$ are a characteristic velocity and length scale respectively.  However, for 2D Stokes flow we have Stokes paradox and slender body effects. The analogous scaling of the stress is
\begin{equation}
\tau \sim \frac{\mu V}{L \log(W/L)}
\end{equation}
where $W$ is the width in the third dimension. In the following, we will take $W = 2\pi$, since a body cannot have a length greater than $2\pi$ and still fit inside the domain.

\subsection{The vanishing rate of the area}
The rate of area loss is given by
\begin{equation}
\dot{A} = - \abs{\tau} L
\end{equation}
We then insert the scaling $\abs{\tau} \sim -1/(L \log L)$ and simplify to get
\begin{equation}
\dot{A} = C/\log{A}
\end{equation}
which we solve to get
\begin{equation}
A (1 - \log A) = C (t_f - t)
\end{equation}
In particular, I believe the signs are correct here.
BUT, I am not so sure of this calculation anymore, now that $W$ is included.

>>>>

\section{Scaling law and interface smoothing}
For 3D Stokes flow, the shear stress scales like
\begin{equation}
\tau \sim \mu U/a
\end{equation}
where $U$ and $a$ are a characteristic velocity and length scale respectively.  However, for 2D Stokes flow we have Stokes paradox and slender body effects. The analogous scaling of the stress is
\begin{equation}
\tau \sim \frac{\mu U}{a \log(\ell/a)}
\end{equation}
where $\ell$ is the width in the third dimension. 
%For example, we can take $\ell = 2\pi$, since a body cannot have a length greater than $2\pi$ and still fit inside the domain.

\subsection{The vanishing rate of the area}

The area loss rate is given by
\begin{align}
\dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, .
\end{align}
The above is an exact relation. Substitution of the scaling law gives
\begin{align}
\dot{A} \sim \frac{ - \mu U}{\log(\ell/a)} \, .
\end{align}
Inserting $A \sim a^2$ and rearranging gives
\begin{align}
\log (c_2 A/\ell^2) dA = c_3 \mu U dt
\end{align}
Integrating gives
\begin{align}
A \left(1 - \log(c_2 A/\ell^2) \right) = c_3 \mu U (t_f - t)
\end{align}
Rearranging and setting some constants based on knowing $A(t=0) = A_0$ gives
\begin{align}
\label{area_predict}
\frac{A}{A_0} \left( 1 - c_1 \log{\frac{A}{A_0}} \right) \sim \frac{t_f - t}{t_f} \quad \text{as } t \to t_f \, .
\end{align}
where $c_1$ is a constant, and I find $c_1 = 0.24$ gives an excellent fit to the data.

Now differentiation the result with respect to time gives
\begin{align}
\dot{A} = \frac{A_0/t_f}{1 - c_1 (1 + \log(A/A_0))} \sim - \frac{\mu U}{\log(\ell / a)}
\end{align}
Now, doing some manipulations, using that $A = c_3 a^2$ and $A_0 = c_3 a_0^2$, gives the nice relationship
\begin{align}
\ell = a_0 e^{\frac{1-c_1}{2 c_1}} = 4.87 a_0 
\end{align}

Now, to estimate the drag, use the slender-body formula for drag per unit span, $\ell$, which is
\begin{align}
F_D \sim \frac{4 \pi \mu U}{\log(\ell/a)}
\end{align}
The $4 \pi$ coefficient is exact (in the slender-body framework) for a circular cylinder. Substituting the value we found $\ell = 4.87 a_0$, gives
\begin{align}
F_D \sim \frac{4 \pi \mu U}{\log(4.87 a_0/a)}
\end{align}
For non-circular bodies, we can define $a = \sqrt{A/\pi}$ since this is the value that would be consistent with a circle. This gives
\begin{align}
F_D \sim \frac{4 \pi \mu U}{\log(4.87 \sqrt{A_0/A})}
\end{align}
or
\begin{equation}
F_D \sim \frac{8 \pi \mu U}{\log(23.7 A_0/A)}
\end{equation}

Alternatively we can just treat $\ell$ as a fit parameter.

### 2018-03-06 docs/methods/erosion_methods.tex
NM
Methods paper NM
Working on drag. It is kind of subtle to get estimates with slender-body theory.

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress is growing as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / L$, since stress is inversely proportional to body scale $L$. This law must be modified in two dimensions, though.
The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $W \gg L$, the drag is proportional to
$\mu \umax W/ \log(W/L)$. This relationship implies that, for a 2D
cross-section, the stress per unit span can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{L \log(W/L)} \qquad \text{as } L \to 0 \, .
\end{align}
In interpreting our 2D simulations, we will treat the span, $W$, as fixed with the
requirement $W \gg L$.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.85 \textwidth]{./figs/area_drag.pdf}
\caption{The area and drag of an eroding body: (a) The body area measured in the simulations (black) shown against the scaling-law prediction \eqref{area_predict} (dotted red). The two are nearly indistinguishable. (b) The horizontal drag (black) on the body as it erodes, along with the viscous (green) and pressure (blue) contributions. The scaling prediction can be rearranged to give estimate \eqref{viscdrag} for the viscous drag (dotted red).}
\label{area_drag}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-6, fixarea = 0, fixpdrop = 0

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align}
\dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, .
\end{align}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$. Inserting scaling law~\eqref{eqn:stresslaw} gives
\begin{align}
\dot{A} \sim \frac{ - \mu \umax}{\log(W/L)} \quad \text{as } L \to 0 \, . 
\end{align}
Noting that $A \sim L^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align}
\label{area_predict}
A (c_1 - \log A) \sim c_2 (t_f - t) \qquad \text{as } t \to t_f \, .
\end{align}
where $c_1$ and $c_2$ are constants.
This is an {\em implicit} relationship for how the area scales down in time during erosion. In Fig.~\ref{area_drag}a, we show the body area as a function of time measured in the simulation against the above scaling law with $c_1 =2.2$ and $c_2 = 30$. The comparison shows close agreement between the two (they are indistinguishable).

In Fig.~\ref{area_drag}(b), we show the horizontal drag $F_D$ measured on the body as it erodes. Note that, in the context of 3D slender-body theory, $F_D$ is drag per unit span. The figure shows the drag to decrease  gradually at first, then abruptly tend to zero in the final moments of the body vanishing. This behavior is consistent with fact that drag vanishes {\em logarithmically} with $L$  in 2D Stokes flow. We also show in the figure the separate contributions of the viscous and the pressure drag. Initially, these two contributions are comparable, with a slightly greater level of pressure drag. However, as the body erodes and assumes a more slender form, the pressure drag vanishes much more rapidly than viscous drag.

As a final check, are scaling estimates for area change can be rearranged to estimate the viscous drag component
\begin{align}
\label{viscdrag}
F_{\mu} \approx \mean{\atau} L \sim \frac{c_2}{-\log A + c_1 -1}
\end{align}
This scaling law is shown in the figure (red, dotted curve) and shows good agreement with the viscous drag. Of course, \eqref{viscdrag} is less exact than the previous estimates since it does not take into account the normal direction of the surface, only the mean shear stress.

>>>>

Figure~\ref{shrink_intface}(a) shows that the spacing between successive interfaces becomes larger with time, suggesting that the overall shear stress is growing as the body vanishes. This observation is consistent with the well-known 3D Stokes drag law, $\tau \sim \mu \umax / a$, in which drag is inversely proportional to body-scale $a$ (e.g.~for a sphere, $a$ is taken to be the radius). This law must be modified in two dimensions, though. The modification is made non-trivial by subtleties of the 2D Stokes paradox, but can be correctly deduced by noting that, as the body vanishes, its drag must be consistent with slender-body theory \cite{poz1997, bouzarth2007epicyclic, camassa2008trajectory, MooreJFM2012, zhao2015transportation}. 
For a slender body in flow perpendicular to its long-axis $\ell$, the drag is proportional to
$\mu \umax \ell/ \log(\ell/a)$, where $a$ is a lengthscale of the cross-section and $a \ll \ell$. This relationship implies that the stress per unit span on a 2D cross-section can be estimated by
\begin{align}
  \label{eqn:stresslaw}
  \tau \sim \frac{\mu \umax}{a \log(\ell/a)} \qquad \text{for } a \ll \ell \, .
\end{align}
In interpreting our 2D simulations, we will treat the span, $\ell$, as fixed. Recall that $L$ denotes the total perimeter of the cross-section, and therefore $a$ and $L$ can often be used interchangeable in scaling analysis, since they each represent a lengthscale of the cross-section.

%^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
\begin{figure}%[htbp]
\begin{center}
\includegraphics[width = 0.9 \textwidth]{./figs/area_drag.pdf}
\caption{The area and drag of an eroding body: (a) The body area measured in the simulations (black) shown against the scaling-law prediction \eqref{area_predict} (dotted red). The two are nearly indistinguishable. (b) The horizontal drag (black) on the body as it erodes, along with the viscous (green) and pressure (blue) contributions. The scaling prediction can be rearranged to give estimate \eqref{viscdrag} for the viscous drag (dotted red).}
\label{area_drag}
\end{center}
\end{figure}
 %^^^^^^^^^^^^^^^^^^^^^^^^^^^^^^%
% Figure from datafile 01circ1024aaa;
% Uses 01circ1024.in with epsfac = 10, sigfac = 10, dt = 1e-7, fixarea = 0, fixpdrop = 0

The body's rate of area reduction is given by integrating the absolute shear stress around the boundary
\begin{align}
\dot{A} = - \int_{\gamma} \atau \, ds \,  = - \mean{\atau} L \, .
\end{align}
Thus, the area-loss rate is given by the product of the mean shear
$\mean{\atau}$ and the total perimeter $L$, which is an exact relationship. Inserting scaling law~\eqref{eqn:stresslaw} gives
\begin{align}
\dot{A} \sim \frac{ - \mu \umax}{\log(\ell/a)} \qquad \text{as } a \to 0 \, . 
\end{align}
Noting that $A \sim a^2$ allows one to solve the ODE to determine how the body area changes with time
\begin{align}
\label{area_predict}
\frac{A}{A_0} \left( 1 - c_1 \log{\frac{A}{A_0}} \right) \sim \frac{t_f - t}{t_f} \quad \text{as } t \to t_f \, .
\end{align}
Here, $t_f$ is the time of vanishing, $A_0$ is the body's initial area, and $c_1$ is a free constant. The above is an {\em implicit} relationship for how the area scales down in time during erosion. In Fig.~\ref{area_drag}a, we show the body area as a function of time measured in the simulation against the above scaling law with $c_1 = 0.24$ a fit parameter. The comparison shows close agreement between the two (they are indistinguishable).

In Fig.~\ref{area_drag}(b), we show the horizontal drag $F_D$ measured on the body as it erodes. Note that, in the context of 3D slender-body theory, $F_D$ is drag per unit span. The figure shows the drag to decrease  gradually at first, then abruptly tend to zero in the final moments of the body vanishing. This behavior is consistent with fact that drag vanishes {\em logarithmically} with body-size $a$  in 2D Stokes flow. We also show in the figure the separate contributions of the viscous and the pressure drag. Initially, these two contributions are comparable, with a slightly greater level of pressure drag. However, as the body erodes and assumes a more slender form, the pressure drag vanishes much more rapidly than viscous drag.

As a final check, we can estimate the drag (per unit span) using slender-body theory via the formula
\begin{align}
F_D \sim \frac{4 \pi \mu U}{\log(\ell/a)}
\end{align}
For a circular cylinder, $a$ is the radius and the prefactor $4 \pi$ above gives an exact result within the slender-body framework. To generalize to other shapes, we define $a = \sqrt{A/\pi}$ and fit $\ell$ to get $\ell = 2.3 a_0$.

\todo[inline]{Clean up this section}
